A Practical Optional Type System for Clojure

### 2012-10-24 dissertation/intro.tex
Filling in the last bits. Mostly design and future work

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This is idea not new, but recent attempts are noteworthy for 
their broad success in matching many of the advantages of statically typed
languages, notably due to the use of \emph{bidirectional checking}.
Instead of always attempting to infer types, this algorithm relies on programmer annotations
appearing in some natural places such as giving the type of each top-level function.

The Clojure programming language is a dynamically typed dialect of Lisp invented
by Hickey, designed to run on popular platforms\footnote{http://clojure.org/}.
It emphasises functional programming with immutable data structures
and provides direct interoperability with its host platform.
Notable implementations of Clojure exist for the Java Virtual Machine (JVM),
the Common Language Runtime, and for Javascript virtual machines.
At the current time, Clojure on the JVM is the most mature implementation,
and therefore this project focuses on the JVM implementation.

@@

The initial motivation for implementing an optional type system
for Clojure was an anecdotal account of the struggles
of a Clojure programmer. In an apparently heroic effort, he managed to 
implement a Clojure library for conduits, an advanced form of ``pipes'',
using arrows, a generalisation of monads.
\footnote{CONDUIT REFERENCE TODO}
Conduits also significantly surpass monads in complexity and are usually reserved
for languages with advanced type systems like Haskell.

@@

Since then, the Clojure community has shown significant interest in this work.
I also developed this project as a Google Summer of Code 2012 project,
after it was selected by the Clojure community as a Clojure Google Summer of Code
project for 2012.
\footnote{CITE GSOC, or FOOTNOTE}
I have also been invited to speak on this project at the Clojure Conj
2012 conference in November, the main international conference related to Clojure.

@@

% refinements check invariants of programs
% while ordinary static types check whether programs
% are basically meaningful - Rowan + Frank P
% Page 1 Rowan's PhD

There are many concepts associated with \emph{types} and \emph{type systems} in both the
literature and informal discourse.
A programmer who uses dynamically-typed languages may have a drastically different notion
of what a type is than, say, a programmer preferring languages with advanced static type systems.
There is some debate as to whether optional static type systems like Typed Clojure
can even be called a type system. We choose to
follow the terminology of Pfenning~\cite{Pfe08} and Reynolds~\cite{Rey01},
where such optional type systems are \emph{extrinsic} type systems, and more
traditional type systems are \emph{intrinsic} type systems.
This distinction has a long history, originating in work in the $\lambda$-calculus 
by Church \footnote{CHURCH FROM ROWANS DISSERTATION} and Curry \footnote{CURRY FROM ROWANS DISSERTATION},
leading to intrinsic types and extrinsic types also being called Church types and
Curry types often in the literature.

@@

@@

\subsection{Datatypes and Protocols}

As well as \lstinline|def| and \lstinline|defn| definitions,
Clojure programmers typically include datatype and protocol
definitions. 
Protocols are similar to Java interfaces\footnote{JAVA INTERFACE REFERENCE} and datatypes to classes implementing interfaces.
We can annotate datatype and protocol definitions similarly.

>>>>

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This is idea not new~\cite{Fag91,CF91,AM91}, but recent attempts~\cite{Tob10}
are noteworthy for their broad success in matching many of the advantages of statically typed
languages, notably due to the use of bidirectional checking~\cite{PT00}.
Instead of always attempting to infer types, this algorithm relies on programmer annotations
appearing in some natural places such as giving the type of each top-level function.

The Clojure programming language is a dynamically typed dialect of Lisp invented
by Hickey \footnote{\url{http://clojure.org/}}, designed to run on popular platforms.
It emphasises functional programming with immutable data structures
and provides direct interoperability with its host platform.
Notable implementations of Clojure exist for the Java Virtual Machine (JVM),
the Common Language Runtime, and for Javascript virtual machines.
At the current time, Clojure on the JVM is the most mature implementation,
and therefore this project focuses on the JVM implementation.

@@

The initial motivation for implementing an optional type system
for Clojure was an anecdotal account of the struggles
of a Clojure programmer. In an apparently heroic effort, he managed to 
implement a Clojure library for conduits~\footnote{\url{http://www.yesodweb.com/book/conduits}}, 
an advanced form of ``pipes'', using arrows, a generalisation of monads.
Conduits also significantly surpass monads in complexity and are usually reserved
for languages with advanced type systems like Haskell.

@@

Since then, the Clojure community has shown significant interest in this work.
I also developed this project as a Google Summer of Code 2012 project,
after it was selected by the Clojure community as a Clojure Google Summer of Code
project for 2012~\footnote{\url{http://www.google-melange.com/gsoc/project/google/gsoc2012/ambrosebs/10001}}.
I have also been invited to speak on this project at the Clojure Conj
2012 conference in November, the main international conference related to Clojure.

@@

% Page 1 Rowan's PhD

There are many concepts associated with \emph{types} and \emph{type systems} in both the
literature and informal discourse.
A programmer who uses dynamically-typed languages may have a drastically different notion
of what a type is than, say, a programmer preferring languages with advanced static type systems.
There is some debate as to whether optional static type systems like Typed Clojure
can even be called a type system. We choose to
follow the terminology of Pfenning~\cite{Pfe08} and Reynolds~\cite{Rey01},
where such optional type systems are \emph{extrinsic} type systems, and more
traditional type systems are \emph{intrinsic} type systems.
This distinction has a long history, originating in work in the $\lambda$-calculus 
by Church~\cite{Chu40} and Curry~\cite{Cur34},
leading to intrinsic types and extrinsic types also being called Church types and
Curry types often in the literature.

@@

\section{Contributions}

In this dissertation we develop an optional type system for Clojure.
We focus on the features needed to make a practical type system that
existing Clojure programmers can use.
We also developed a prototype based on Typed Racket~\cite{Tob10}.

The contributions of the work described in this dissertation are as follows:

\begin{itemize}
  \item We describe a novel use of occurrence typing~\cite{TF10} for eliminating
        occurrences of Java's \emph{null}. We make \emph{null} directly expressible
        in Typed Clojure's type system as \lstinline|nil|. This makes it reasonable for it to be a
        static type error if \emph{null} is potentially misused, an improvement over Java's static type system
        where \emph{null} cannot be expressed.

  \item We show how a combination of intersection types and occurrence typing enables us to accurately type check
        the most common usages of Clojure's sequence abstraction. This was particularly subtle, as 
        many base operations in the sequence abstraction take and/or return \lstinline|nil|, and
        each case could imply different information about a sequence.

  \item We develop a prototype type checker for Clojure running on the Java Virtual Machine
        which is able to type check many common Clojure idioms.
        We show how this type checker can be used to aid Clojure programmers use sophisticated
        programming styles by porting most of an existing library for monadic programming.

  \item This prototype is extended to support several features that are not present in Typed Racket.
        We support filtering sequences on anonymous functions that convey only negative type information
        (eg. non-nil), capturing a common Clojure idiom.

  \item We capture the Clojure idiom of using maps with known keyword fields by using heterogeneous map types.
        These types are sufficient for accurately associating and dissociating known keyword keys,
        and accurately looking up values associated with keyword keys in heterogeneous map types that have
        that keyword key present.
        This is sufficient for type checking many common idioms involving maps.

  \item We identify major future issues in typing Clojure code. Multimethod definitions are problematic
        to type check accurately because of the interaction of its dispatch mechanism and occurrence typing.
        Records have subtle semantics for dissociating keys, like dissociating a base key of a record returns
        a plain map.
%        Heterogeneous map types 
\end{itemize}

Together,

These contributions support my thesis by using the prototype to check practical code
and

@@

\subsection{Datatypes and Protocols}
\label{sec:datatypes}

As well as \lstinline|def| and \lstinline|defn| definitions,
Clojure programmers typically include datatype and protocol
definitions. 
Protocols are similar to Java interfaces~\cite{InterfaceTutorial} and datatypes are similar to classes implementing interfaces.
We can annotate datatype and protocol definitions similarly.

### 2012-10-24 dissertation/lit-review.tex
Filling in the last bits. Mostly design and future work

\section{Static Type Systems}

% Haskell
% OCaml
% Forsythe
% Java C#
% Scala

% cite: Dynamic Typing in Polymorphic languages (1991?)
% - coming from the other direction
% - extending static languages to be more dynamic

% A more recent reference in the same area is "Typed Open Programming" a PhD dissertation from 2007 - at least have a look at 1.3.2 for a good summary of work on dynamics in statically typed languages since 1991 - probably it's good to cite this as a good recent survey of the area.
\section{Static Types for Untyped Languages}

@@

This section summarises research in typing OCaml-style records
which relate to Clojure records and heterogeneous hash maps.

@@

% Hiyashi Singleton types
% Intersection types - CDV81
% Unions in soft typing CF91
% - Static Type Inference in a Dynamically Typed Lanugage
%  - cited by CF91, which has unions

@@

Intersection and union types are interesting type constructs relevant to capturing the complicated
types common in dynamic languages.
An expression of type \lstinline|(I a b)|, an intersection type including types \lstinline|a|
and \lstinline|b| in Typed Clojure, can be used safely in both positions expecting type \lstinline|a|,
and positions expecting type \lstinline|b|.
An expression of type \lstinline|(U a b)|, a union type including types \lstinline|a| in Typed Clojure,
and expressions of this type can be used safely in positions 
that expect a type that is \emph{either} type \lstinline|a|
or \lstinline|b|.
For example, \lstinline|(U Number Boolean)| cannot be used in positions expecting \lstinline|Number|.
% TODO Intersection example

Several interesting projects have used intersection or union types.

Forsythe, a modern ALGOL dialect by Reynolds~\cite{Rey96} was the first wide-spread 
programming language to use intersection types.
Uses of intersections in Forsythe include representing extensible record types
and function overloading.

Refinement types add a level of types refining an existing type system
in order to type check more detailed properties and invariants than standard static type systems
(described by Freeman and Pfenning~\cite{FP91}).
Intersection types are critical here to allow more than one property or invariant
to be expressed for a function.
SML CIDRE is a refinement type checker for Standard ML by Davies~\cite{Dav05}.
\begin{verbatim}
Refinement types are similar to this project ... how?
\end{verbatim}

St-Amour, Tobin-Hochstadt, Flatt, and Felleisen
describe the \emph{ordered intersection types} used in Typed Racket~\cite{St12}
that provide a kind of function overloading.
Typed Clojure takes a similar approach
\begin{verbatim}
described in chapter
\end{verbatim}
Typed Racket also uses disjoint union types,

@@

Scala and Clojure are both languages with a focus on host interoperability, specifically
to the Java Virtual Machine. Their differing treatment of the Java Virtual Machine's \lstinline|null| is
relevant, particularly considering the potential of using occurrence typing~\cite{TF10}
to prevent erroneous usage of \lstinline|null|.
\begin{verbatim}
eg.
\end{verbatim}

@@

>>>>

\section{Dynamic Typing in Typed Languages}

This section 
\cite{ACPR94}

% cite: Dynamic Typing in Polymorphic languages (1994)
% - coming from the other direction
% - extending static languages to be more dynamic

% A more recent reference in the same area is "Typed Open Programming" a PhD dissertation from 2007 - at least have a look at 1.3.2 for a good summary of work on dynamics in statically typed languages since 1991 - probably it's good to cite this as a good recent survey of the area.
\section{Static Typing for Untyped Languages}

@@

OCaml-style extensible record types has been the subject of extensive research~\cite{Wan89,CM91,HP91}.
Typed Clojure's heterogeneous map types show some resemblance to extensible
record types, but a survey of the literature did not reveal research flexible
enough to capture common usages of maps in Clojure.
This most likely implies that designing such a system for Typed Clojure would be hard.
For this reason, heterogeneous map types are kept as simple as possible while still being useful.

@@

% Hayashi Singleton types
% Intersection types - CDV81
% Unions in soft typing CF91
% - Static Type Inference in a Dynamically Typed Lanugage
%  - cited by CF91, which has unions

@@

Intersection and union types are interesting type constructs relevant to capturing the complicated
types common in dynamic languages.
An expression of type \lstinline|(I a b)|, an intersection type including types \lstinline|a|
and \lstinline|b| in Typed Clojure, can be used safely in both positions expecting type \lstinline|a|,
and positions expecting type \lstinline|b|.
An expression of type \lstinline|(U a b)|, a union type including types \lstinline|a| in Typed Clojure,
and expressions of this type can be used safely in positions 
that expect a type that is \emph{either} type \lstinline|a|
or \lstinline|b|.
For example, \lstinline|(U Number Integer)| cannot be used in positions expecting \lstinline|Integer|,
but \lstinline|(I Number Integer)| can be used in positions expecting \lstinline|Integer|.

Intersection, union, and singleton types fit best as extrinsic types.
Hayashi~\cite{Hay91} describes two type theories, ATT which includes
intersection, union, and singleton types, and ATTT which further extends ATT
to include refinement types, which can be classified as extrinsic types.
Typed Clojure can also be considered an extrinsic type system which supports
union, intersection, and singleton types, but in less sophisticated forms
than ATT and ATTT.
In particular, singleton types in ATTT a much more advanced than Typed Clojure,
which offers only singleton types for values like \lstinline|:a| and \lstinline|nil|
and, unlike ATTT, no way to parameterise over singletons.

Several interesting projects have used intersection or union types.
Intersection types were originally introduced by Coppo, Dezani-Ciancaglini, and Venneri~\cite{CDV81}.
The use union types to type check dynamically typed languages with soft typing systems
dates back to Cartwright and Fagan~\cite{CF91}.

Forsythe, a modern ALGOL dialect by Reynolds~\cite{Rey81,Rey96} was the first wide-spread 
programming language to use intersection types.
Uses of intersections in Forsythe include representing extensible record types
and function overloading.

Refinement types add a level of types refining an existing type system
in order to type check more detailed properties and invariants than standard static type systems
(described by Freeman and Pfenning~\cite{FP91}).
Refinement types are reminiscent of Typed Clojure, in that Typed Clojure
adds a level of types on top of Clojure.
Intersection types are critical here to allow more than one property or invariant
to be expressed for a function.
SML CIDRE is a refinement type checker for Standard ML by Davies~\cite{Dav05}.

St-Amour, Tobin-Hochstadt, Flatt, and Felleisen
describe the \emph{ordered intersection types} used in Typed Racket~\cite{St12}
that provide a kind of function overloading.
Typed Clojure takes a similar approach for its representation of functions.
Typed Racket uses (disjoint) union types.

@@

Scala and Clojure are both languages with a focus on host interoperability, specifically
to the Java Virtual Machine. Their differing treatment of the Java Virtual Machine's \lstinline|null| is
relevant, particularly considering the potential of using occurrence typing~\cite{TF10}
to prevent erroneous usage of \lstinline|null|.

@@

SML CIDRE \cite{Dav05} was also a secondary influence - particularly it's use of bidirectional 
checking of extrinsic types including subtyping, intersection types, and it's aim to capture program invariants.  
However, SML CIDRE starts from the intrinsically typed language SML which already has a rich static 
type system, which leads to many different considerations compared to starting from a dynamically typed language.  
Much of this influence was via a student-supervisor relationship, making it particularly hard to pin down the specifics.

### 2012-10-24 dissertation/paper.tex
All of Rowan's corrections
Filling in the last bits. Mostly design and future work

\usepackage{mmm}
\usepackage{mathpartir}
\usepackage{clj-grammar}
\usepackage[style=alphabetic]{biblatex}
\addbibresource{bibliography.bib}

@@

Traditionally there has been a clear distinction between 
programming languages that perform type checking at compile-time, and those that do not.
In particular, the early wide-spread programming languages ALGOL and LISP
could be divided this way: ALGOL performs compile-time (or static) type checking,
and LISP does not.
Static type checking eliminates many common user errors that are otherwise unnoticed or are caught 
when the program is run.
On the other hand, a language without static type checking can often be designed 
to be more flexible and forgiving.

A recent trend is to combine the advantages of both kinds of languages by adding optional static 
type systems to languages without static type checking. 
A major challenge in these combinations is continuing to support the idiomatic style of the original 
language, and some recent attempts appear to have finally produced type systems that have overcome this obstacle.  
Most notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
check most idiomatic Racket code.

@@

This paper describes my work on designing an optional static type system for Clojure based on
the lessons learnt from several projects, primarily Typed Racket.
A prototype type system is developed for Clojure running on the Java Virtual Machine.
Several experiments are conducted using this prototype, particularly involving
sufficiently complicated existing Clojure code where type checking increases confidence that the code correct.
For example, a significant portion of algo.monads, a Clojure Contrib library for monadic programming, 
is able to be type checked. 
On the positive side, most monad, monad transformer, and monadic function definitions can be type checked,
usually by adding type annotations in natural places like function definitions.
Negatively, simple monad comprehensions can be type checked, but require awkward type annotations.

There is significant future work to fully type check all Clojure features and idioms.
For example, multimethod definitions and functions accepting an even number of variable arguments
are troublesome. 
Also, there are desirable features from the Typed Racket project that are missing, such
as automatic runtime contract generation and a sophisticated blame system, 
both which are designed to improve error messages when mixing typed and untyped code in similar systems.
I conclude that it appears to be both practical and useful to design and implement an optional static type system for the
Clojure programming language.

@@

% Evaluation
% - successes
% - future work identified

@@

%
%\begin{mathpar}
%\inferrule*[right=B-Var]
%  {\rho(x) = v}
%  {\rho \vdash x \downarrow v}
%
%\inferrule*[right=B-Assoc]
%  {\rho \vdash e_m \downarrow mi \\
%   \rho \vdash e_k \downarrow k \\
%   \rho \vdash e_v \downarrow v \\
%   \rho \vdash \overrightarrow{e_j}^n \downarrow \overrightarrow{k_j}^n \\
%   \rho \vdash \overrightarrow{e_m}^n \downarrow \overrightarrow{k_m}^n \\
%   \delta(add\_entry, mi, k, v, \overrightarrow{k_j, k_m}^n) = mo}
%  {\rho \vdash (assoc\ e_m\ e_k\ e_v\ \overrightarrow{e_j\ e_m}^n) \downarrow mo}

%\infrule[B-Var]
%  {\rho(x) = v}
%  {\rho \vdash x \downarrow v}
%
%\infrule[B-Assoc]

%\end{mathpar}

@@

@@

>>>>

\usepackage{mmm}
\usepackage{mathpartir}
\usepackage{clj-grammar}
\usepackage{url}
\usepackage[style=alphabetic]{biblatex}
\addbibresource{bibliography.bib}

@@

Traditionally there has been a clear distinction between 
programming languages that perform type checking at compile-time, and those that do not.
In particular, the early wide-spread programming languages ALGOL and Lisp
could be divided this way: ALGOL performs compile-time (or static) type checking,
and Lisp does not, and most recent languages can be similarly classified.
Static type checking eliminates many common user errors that are otherwise unnoticed or are caught 
when the program is run.
On the other hand, a language without static type checking can often be designed 
to be more flexible and forgiving, making it easy, for example, to have a \emph{map} function that can map functions
with any number of arguments (as in most Lisp dialects).

A recent trend is to aim to combine the advantages of both kinds of languages by adding optional static 
type systems to languages without static type checking. 
A major challenge in these combinations is continuing to support the idiomatic style of the original 
language while performing accurate enough checking to catch similar errors to standard statically
typed languages. Some recent attempts appear to have finally produced type systems that have overcome this obstacle.
Notably Typed Racket, a typed sister language of the Lisp dialect Racket, is able to statically type 
check most idiomatic Racket code.
These optional type systems have some important differences from standard static
type systems, for example they often include \emph{union} types.

@@

This dissertation describes my work on designing an optional static type system for Clojure based on
the lessons learnt from several projects, primarily Typed Racket.
This work includes designing and building a type checker for Clojure running on the Java Virtual Machine.
Several experiments are conducted using this prototype, particularly involving
existing Clojure code that is sufficiently complicated that
type checking increases confidence that the code is correct.
For example, nearly all of \emph{algo.monads}, a Clojure Contrib library for monadic programming, 
is able to be type checked.
Most monad, monad transformer, and monadic function definitions can be type checked,
usually by adding type annotations in natural places like function definitions.

There is significant future work to fully type check all Clojure features and idioms.
For example, multimethod definitions and functions accepting an even number of variable arguments
are troublesome. 
Also, there are desirable features from the Typed Racket project that are missing, such
as automatic runtime contract generation and a sophisticated blame system, 
both which are designed to improve error messages when mixing typed and untyped code in similar systems.
Overall, the work described in this dissertation leads to the conclusion that it appears to 
be both practical and useful to design and implement an optional static type system for the
Clojure programming language.

@@

@@

@@

Whether a language is dynamically typed or statically typed is starting to
blur. Recent languages have managed to support the advantages of both styles:
the safety of a statically typed language while retaining the idioms found in
dynamically typed languages.

Extensions like Typed Clojure

@@

\include{appendix}
